### 2020-12-08 submission/manuscript.Rmd
First draft of Intro/Abstract/Importance sections

\newpage
<!-- \linenumbers -->

@@

250 words

@@

## Importance
150 words

@@

Two elements that are held in tension in the analysis of 16S rRNA gene sequence data is how to adequately remove PCR and sequencing artifacts while decreasing the granularity of the taxonomic level that is used in the analysis. Previous attempts have included screening for sequencing quality based on quality scores [Kozich/Edgar] followed by a polishing step based on the frequency of the sequences relative to similar sequences [Kozich/Edgar/DeBlur]. Other pipelines model the quality scores and types of errors to cluster sequences directly [Dada2]. But, as a final step many pipelines advocate for removing rare sequences from each dataset prior to outputting the sequence data as amplicon sequence variants (ASVs) [Knight/Edgar/DeBlur/Dada2]. ASVs are often clustered further to generate operational taxonomic units or phylotypes. Some pipelines remove all ASVs that appear once (i.e. singletons) [XXXX], XXXXXX [XXXXXX], XXXXX [XXXXXXX], or XXXXXX [XXXXXX] times prior to further clustering or making ecological comparisions. Notably, the mothur-based pipeline discourages the practice of removing rare sequences.

The abundance-based screening approach assumes that rare ASVs are more likely to be artifacts than more abundant ASVs. Sequencing of mock communities confirms that artifacts tend to be rare. Proponents of abundance-based screening point to their ability to obtain the correct number of ASVs, OTUs, or phylotypes with data generated from sequencing mock communities when rare ASVs are removed. However, this approach effectively overfits the curation pipeline to data generated from a phylogenetically simple community with an atypical community distribution that is often sequenced to a depth that is not achieved with biological samples. It is necessary to think more deeply about the practice of abundance-based screening.

The minimum abundance thresholds that have been proscribed are developed and applied without regard for the total number of sequences generated from each sample. Ignored in their recommendations is the common experience that the number of sequences generated from each sample may vary by two or three orders of magnitude. An ASV that appears once in a sample with 2,000 sequences is more trustworthy than an ASV that appears once in a sample with 100,000 sequences since it has a 50-fold higher relative abundance. But, according to the pipeline recommendations, they are treated as being equally trustworthy. Rather than removing rare ASVs, the approach taken by the mothur pipeline applies the classical ecological approach of rarefaction. Each sample is rarefied to the same sequencing depth so that the number of artifacts that appears in each sample is controlled.

Experience sequencing biological samples demonstrates that there are good ASVs that may have an abundance below the proscribed threshold. For example, the abundance of an ASV may be below the threshold in some samples or time points and above the threshold in others. However, rarity, both in terms of prevalence and incidence, is an important ecological concept. Removing rare ASVs likely hinders one's ability to ability to make inferences about the dynamics and nature of the populations that rare ASVs represent. Furthermore, removing ASVs whose abundances are below the proscribed threshold also potentially biases the community structure of the samples.

In the current study, I use published sequence data from `r nrow(table_1)` studies to investigate the nature of rare ASVs (i.e. those that appear 10 or fewer times) and the effects of removing them on downstream analysis of microbial communities. The analysis was also performed using traditional operational taxonomic units, where ASVs subjected to abundance-based screening were clustered such that the ASVs within an OTU were no more than 3% different from each other. The results reject the assumptions built into abundance-based screening and highlight the problems inherent in removing rare ASVs.

>>>>

<!-- \linenumbers -->

@@

Methods for remediating PCR and sequencing artifacts in 16S rRNA gene sequence collections are in continuous development. The choices that researchers make have significant ramifications on the inferences that can be drawn from microbial ecology studies. One common approach is to remove rare sequences from datasets. But, the threshold that is applied is generally selected without regard for the number of sequences in the sample or the variation in sequencing depth across samples within a study. I analyzed the impact of removing rare sequences on metrics of alpha and beta diversity using samples collected across `r nrow(table_1)` published datasets. Removal of rare sequences significantly decreased the number of amplicon sequence variants and operational taxonomic units as well as their diversity. Furthermore, their removal increased the variation in community structure between samples. When a known effect size was simulated, removal of rare sequences reduced the statistical power to detect the effect relative to not removing rare sequences. Although removal of rare sequences did not affect the probability of falsely detetecting differences when samples were randomized to simulate a null model, the false detection rate increased when samples were assigned to simulated treatment groups according to their sequencing depth and rare sequences were removed. The false detection rate did not vary when rare sequences were retained. This analysis demonstrates the problems inherent in removing rare sequences. Researchers are encouraged to retain these sequences, to select approaches that minimize PCR and sequencing artifacts, and to use rarefaction to control for uneven sampling effort.

@@

## Importance

Removing rare sequences from 16S rRNA gene sequence collections is an approach that has grown in popularity for handling PCR and sequencing artifacts. Yet, it is unclear what impact an abundance-based filter has on downstream analyses. To investigate the effects of removing rare sequences, I analyzed the community distributions found in the samples of `r nrow(table_1)` published datasets. Analysis of these data and simulations based on them showed that removal of rare sequences distorts the representation of microbial communities. This has the effect of artificially making it more difficult to detect differences between treatment groups. Also of concern was the observation that if sequencing depth is confounded with treatment group, then the probability of falsely detecting a difference between the treatment groups increases with the removal of rare sequences. The practice of removing rare sequences should stop, lest researcher adversely affect the interpretation of their data.

@@

## Introduction

16S rRNA gene sequencing is a mainstay of microbial community analysis [@Knight2018]. Two elements that are held in tension in the analysis of 16S rRNA gene sequence data are how to adequately remove PCR and sequencing artifacts while decreasing the granularity of the taxonomic level that is used in the analysis. When coarse taxonomic levels (e.g. genus level) are used, the effects of artifacts are minimized since the genetic breadth of the level is wider than the diversity of artifacts. Conversely, with fine taxonomic levels (e.g. amplicon sequence variants; ASVs) the effects of artifacts are significant since each artifact may represent a new ASV.

Numerous studies have attempted to address the problem of removing or "denoising" artifacts from data generated using Illumina's MiSeq platform. In one approach, paired sequence reads are aligned and any discrepencies between the reads are resolved based on the difference in quality score for the position in question [@Kozich2013; @Edgar2016; @Masella2012]; quality scores are also used to curate single reads [@Bokulich2013]. In addition, a polishing step is often used to identify ideal sequecnces and their related artifacts based on the frequency of the sequences and their similarity sequences [@Kozich2013; @Edgar2016; @Amir2017]. In a second approach, the quality scores and types of errors to cluster sequences directly [@Callahan2016]. Regardless of the approach, many pipelines advocate for abundance-based sceeening where rare sequences are removed from each dataset prior to outputting the sequence data as ASVs [@Edgar2016; @Amir2017; @Callahan2016]. Some algorithms recommend removing all ASVs that appear one (i.e. singletons) [@Callahan2016], eight [@Edgar2016], or ten [@Amir2017] or fewer times; these pipelines also vary in whether the minimum abundance threshold should be applied to individual samples [@Callahan2016; @Amir2017] or the pool of samples in a study [@Edgar2016]. A notable exception, the mothur-based pipeline discourages the practice of removing rare sequences [@Kozich2013]. After assigning sequences to ASVs, they are often analyzed as a taxonomic unit or clustered to generate operational taxonomic units (OTUs) or phylotypes.

The abundance-based screening approach assumes that rare ASVs are more likely to be artifacts than more abundant ASVs. Sequencing of mock communities confirms that artifacts tend to be rare [@Kozich2013; @Bokulich2013]. Proponents of abundance-based screening point to their ability to obtain the correct number of ASVs, OTUs, or phylotypes with data generated from sequencing mock communities when rare ASVs are removed [@Bokulich2013]. However, this approach effectively overfits the curation pipeline to data generated from a phylogenetically simple community with an atypical community distribution that is often sequenced to a coverage that is not achieved with biological samples. It is necessary to think more deeply about the practice of abundance-based screening.

The minimum abundance thresholds that have been proscribed were developed and applied without regard for the number of sequences generated from each sample.  These recommendations appear to assume that the number of reads per sample is consistent within and between studies. However, it is common that the number of sequences generated from each sample may vary by two or three orders of magnitude (e.g. Figure S1). An ASV that appears once in a sample with 2,000 sequences is more trustworthy than an ASV that appears once in a sample with 100,000 sequences since it has a 50-fold higher relative abundance. But, according to the pipeline recommendations, they are treated as being equally trustworthy. Rather than removing rare ASVs, the approach taken by the mothur pipeline applies the classical ecological approach of rarefaction. Each sample is rarefied to the same sequencing depth so that the number of artifacts that appears in each sample is controlled.

Experience sequencing biological samples also demonstrates that there are bona fide ASVs that may have an abundance below the proscribed threshold. For example, the abundance of an ASV may be below the threshold in some samples or time points and above the threshold in others. However, rarity, both in terms of prevalence and incidence, is an important ecological concept. Removing rare ASVs likely hinders one's to ability to make inferences about the dynamics and nature of the populations that rare ASVs represent. Furthermore, removing ASVs whose abundances are below the proscribed threshold also potentially biases the community structure of the samples.

In the current study, I used published sequence data from `r nrow(table_1)` studies to investigate the nature of rare ASVs (i.e. those that appear 10 or fewer times) and the effect that removing them has on downstream analysis of microbial communities. The analysis was also performed using traditional OTUs, where ASVs subjected to abundance-based screening were clustered such that the ASVs within an OTU were no more than 3% different from each other. The results reject the assumptions built into abundance-based screening and highlight the problems inherent in removing rare ASVs.
